Towards Practical First-Order Model Counting

### 2024-04-29 doc/paper.tex
Proof of Theorem 1

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

@@

\author{%
  Ananth K. Kidambi$^1$\footnote{The first and second authors contributed equally and were affiliated with the National University of Singapore during the completion of this work.}\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, Mumbai, India\\
$^2$National University of Singapore, Singapore, Singapore\\
$^3$University of Toronto, Toronto, Canada\\
\emails
\{210051002, 210050061\}@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}

@@

\emph{First-order model counting} is the problem of counting the number of
models of a sentence in first-order logic over some given domain(s). The
(symmetric) weighted variant of the problem (WFOMC) asks for the sum of model
weights instead, where the weight of a model is a product over predicate
weights~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC is a prominent approach to
the broader problem known as \emph{lifted (probabilistic) inference} that aims
to compute probabilities more efficiently by utlilising symmetries inherent in
the description of the problem~\cite{DBLP:conf/ecai/Kersting12}.

@@

Lifted inference is an active area of research with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC itself has been used to perform inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By iteratively considering
domains of increasing sizes, the model count of a formula can be interpreted as
an (integer) sequence. WFOMC algorithms have been used to discover new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} and
conjecture~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
construct~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other recursive
structures that describe these sequences. Recently, WFOMC algorithms have also
been extended and applied to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

@@

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}, i.e., fixing the formula and asking whether there exists an
algorithm that can compute the WFOMC in time polynomial in the domain size(s).
If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}. Beame et
al.~\shortcite{DBLP:conf/pods/BeameBGS15} showed that there exists an unliftable
formula with three variables. It is also known that all formulas with up to two
variables are liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
This liftable fragment of formulas with two variables has been extended with
various axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23},
counting quantifiers~\cite{DBLP:journals/jair/Kuzelka21}, and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

There is a variety of WFOMC algorithms with different underlying principles.
Perhaps the most prominent class of WFOMC algorithms is based on
\emph{first-order knowledge compilation}. Here, by iteratively applying
\emph{compilation rules}, the formula is compiled to a representation (such as a
circuit of graph), which can then be used to compute the WFOMC for any
combination of domain sizes (or weights). Algorithms in this class include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent extension
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles formulas into
circuits whereas the latter compiles them first to \emph{first-order
  computational graphs} (FCGs) and then to (algebraic) equations. Another major
WFOMC algorithm \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} is based on
cell enumeration. Other algorithms employ local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, junction
trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

@@

The recently-proposed \textsc{Crane} algorithm, while able to handle formulas
beyond the reach of \textsc{ForcLift}, was incomplete in that it could only
construct functions definitions, which would then need to be evaluated to
compute the WFOMC.\@ Moreover, recursive functions were presented without base
cases. In this work, we present \Cranetwo{}---an extension of \textsc{Crane}
that fixes these two weaknesses.

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas that correspond to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} describes post-processing techniques for FCGs and the
formulas from \cref{step2} above necessary to preserve the correct model count.
Next, \cref{sec:cpp} explains how function definitions that encode a solution to
a WFOMC problem are then compiled to C++ programs.\footnote{Compilation to C++
  has been explored before~\cite{DBLP:conf/kr/KazemiP16}, although with limited
  capabilities in terms of what formulas can be handled in a lifted manner.}
Finally, \cref{sec:experiments} reports on two experiments that compare
\Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{Comparison of three logics used in WFOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item number of sorts,
      \item support for constants,
      \item maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    In this context:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

@@

Both \Ctwo{} and \UFO{} are single-sorted (i.e., variables are always quantified
over the same domain) and only include formulas with at most two variables and
no constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{exactly three positive literals with predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (which
  may not be distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets that, satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is
  considered a \emph{model} if it satisfies $\phi$.
\end{definition}

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Continuing from \cref{def:model}, for each predicate $P$ in $\phi$, let
  $w^{+}(P)$ and $w^{-}(P)$ be its \emph{weights} in $\mathbb{Q}$. Unless
  specified otherwise, we assume weights to be equal to one. The
  \emph{(symmetric) weighted first-order model count} (WFOMC) of $\phi$ (with
  respect to $\sigma$, $w^{+}$, and $w^{-}$) is calculated as follows:
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is taken over all models of $\phi$.
\end{definition}

@@

We write $\expr{}$ to represent an arbitrary algebraic expression. It is
important to note that certain terms have different meanings in the context of
algebra compared to logic. In algebra, a \emph{constant} refers to a
non-negative integer. Likewise, a \emph{variable} can either be a parameter of a
function or a variable introduced through summation, such as $i$ in the
expression $\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is denoted as
$f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short), where $f$ represents an
$n$-ary function, and each $x_{i}$ represents a variable. An \emph{equation} is
written as $f(\mathbf{x}) = \expr{}$, with $f(\mathbf{x})$ representing a
signature.

\begin{definition}[Base case]\label{def:basecase}
  Let $f(\mathbf{x})$ be a function call where each $x_{i}$ is either a constant
  or a variable (note that signatures are included in this definition). Then a
  function call $f(\mathbf{y})$ is considered a \emph{base case} of
  $f(\mathbf{x})$ if $f(\mathbf{y}) = f(\mathbf{x})\sigma$, where $\sigma$ is a
  substitution that replaces one or more $x_{i}$ with a constant.
\end{definition}

@@

\Cref{alg:compilewithbasecases} outlines our overall approach for compiling a
formula into a set of equation that include the required base cases. In short,
we first use the knowledge compilation algorithm of the original \textsc{Crane}
to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure (described in \cref{sec:identifying}) that returns a
set of base cases that we need to find solutions for. For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains
\cref{alg:compilewithbasecases} in more detail.

\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ to be a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  In particular, we examine the first solution that \Cranetwo{} returns for this
  formula.

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Compile. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we get the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, giving us $g(l, 0) = 1$.
\end{example}

Note that these base cases overlap when $l = m = 0$ but are consistent with each
other since $0^{0} = 1$. More generally, let $\phi$ be a formula with two
domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the model
count of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same
as the model count of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

Finally, we note that the \Simplify procedure plays a crucial role in
simplifying a common algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$.
Here:
\begin{enumerate*}[label=(\roman*)]
  \item $n$ is a variable,
  \item $a, b \in \mathbb{N}_{0}$ are constants,
  \item $f$ is an expression that may depend on $m$, and
  \item $[a \le m \le b] =
  \begin{cases}
    1 & \text{if $a \le m \le b$} \\
    0 & \text{otherwise}
  \end{cases}$ is the Iverson bracket.
\end{enumerate*}
\Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overall}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into the
simpler form above.

@@

\Cref{alg:findbasecases} summarises the implementation of \FindBaseCases. For
each recursive call from a function $f$ to itself, we consider two types of
arguments:
\begin{enumerate*}[label=(\roman*)]
  \item constants and
  \item arguments of the form $x_{i} - c_{i}$, where $c_{i} \in \mathbb{N}_{0}$
  is a constant, and $x_{i}$ is the $i$-th argument of the signature of $f$.
\end{enumerate*}
In the former case, we consider a base case specifically for that constant. In
the latter case, we consider a base case for constants from zero up to (but not
including) $c_{i}$. \Cref{thm:halting} below motivates this approach.

\begin{example}
  Consider the recursive function $g$ from \cref{example:overall}.
  \FindBaseCases{$\mathcal{E}$} iterates over two function calls: $g(l-1, m)$
  and $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the
  latter produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

TODO: Prove (lots of notes, handwritten and elsewhere)

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{fact}{Observation}
\newtheorem{corollary}{Corollary}
\newtheorem{assumption}{Assumption}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

\crefname{fact}{Observation}{Observations}
\crefname{assumption}{Assumption}{Assumptions}

@@

\author{%
  Ananth K. Kidambi$^1$\footnote{The first and second authors contributed equally and were affiliated with the National University of Singapore during the completion of this work.}\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, Mumbai, India\\
$^2$National University of Singapore, Singapore, Singapore\\
$^3$University of Toronto, Toronto, Canada\\
\emails
210051002@iitb.ac.in,
guramrit@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}

@@

\emph{First-order model counting} is the task of counting the number of models
of a sentence in first-order logic over some given domain(s). The (symmetric)
weighted variant of this problem, known as WFOMC, seeks to calculate the sum of
model weights. In WFOMC, the weight of a model is determined by predicate
weights~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC is a key approach to
\emph{lifted (probabilistic) inference}, which aims to compute probabilities
more efficiently by leveraging symmetries inherent in the problem
description~\cite{DBLP:conf/ecai/Kersting12}.

@@

Lifted inference is an active area of research, with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC has been used for inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By considering domains of
increasing sizes, the model count of a formula can be seen as an integer
sequence. WFOMC algorithms have been utilised for discovering new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} as well as
conjecturing~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
constructing~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other
recursive structures that describe these sequences. Additionally, WFOMC
algorithms have been extended to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

@@

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the WFOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}. Beame et
al.~\shortcite{DBLP:conf/pods/BeameBGS15} demonstrated the existence of an
unliftable formula with three variables. It is also known that all formulas with
up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

There are various WFOMC algorithms with different underlying principles. Perhaps
the most prominent class of WFOMC algorithms is based on \emph{first-order
  knowledge compilation}. In this approach, the formula is compiled into a
representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
WFOMC for any combination of domain sizes (or weights). Algorithms in this class
include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles
formulas into circuits, while the latter compiles them first to
\emph{first-order computational graphs} (FCGs) and then to (algebraic)
equations. Another WFOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

@@

The recently proposed \textsc{Crane} algorithm, while capable of handling
formulas beyond the reach of \textsc{ForcLift}, was incomplete as it could only
construct function definitions, which would then need to be evaluated to compute
the WFOMC.\@ Additionally, recursive functions were presented without base
cases. In this work, we introduce \Cranetwo{}, an extension of \textsc{Crane}
that addresses these two weaknesses.

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for FCGs and the
formulas from \cref{step2} above required to preserve the correct model count.
Next, \cref{sec:cpp} elucidates how function definitions encoding a solution to
a WFOMC problem are compiled into C++ programs.\footnote{The possibility of
  compiling WFOMC solutions to C++ programs has been previously
  considered~\cite{DBLP:conf/kr/KazemiP16}. However, the extent of formulas that
  could be handled was limited.} Finally, \cref{sec:experiments} presents two
experiments comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{Comparison of three logics used in WFOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item number of sorts,
      \item support for constants,
      \item maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

@@

\Ctwo{} and \UFO{} are single-sorted (i.e., all variables are quantified over
the same domain) and only include formulas with at most two variables and no
constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{precisely three positive literals with the predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (which
  may not be distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets, satisfying the following conditions:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Continuing from \cref{def:model}, for each predicate $P$ in $\phi$, let
  $w^{+}(P)$ and $w^{-}(P)$ be its \emph{weights} in $\mathbb{Q}$. Unless
  specified otherwise, we assume all weights are equal to one. The
  \emph{(symmetric) weighted first-order model count} (WFOMC) of $\phi$ (with
  respect to $\sigma$, $w^{+}$, and $w^{-}$) is:
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

@@

We write $\expr{}$ to represent an arbitrary algebraic expression. It is
important to note that some terms have different meanings in algebra compared to
logic. In algebra, a \emph{constant} refers to a non-negative integer. Likewise,
a \emph{variable} can either be a parameter of a function or a variable
introduced through summation, such as $i$ in the expression
$\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short), where $f$ represents an
$n$-ary function, and each $x_{i}$ represents a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, with $f(\mathbf{x})$ representing a signature.

\begin{definition}[Base case]\label{def:basecase}
  Let $f(\mathbf{x})$ be a function call where each $x_{i}$ is either a constant
  or a variable (note that signatures are included in this definition). Then
  function call $f(\mathbf{y})$ is considered a \emph{base case} of
  $f(\mathbf{x})$ if $f(\mathbf{y}) = f(\mathbf{x})\sigma$, where $\sigma$ is a
  substitution that replaces one or more $x_{i}$ with a constant.
\end{definition}

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we
utilise the knowledge compilation algorithm from the original \textsc{Crane} to
compile the formula into three components:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is then passed to the
\FindBaseCases procedure (explained in \cref{sec:identifying}), which identifies
the base cases that need to be defined. For each base case $f(\mathbf{x})$, we
determine the formula associated with the function name $f$ and simplify it
using the \Propagate procedure (explained in \cref{sec:simplifying}). The
algorithm then calls itself on these simplified formulas and adds the resulting
base case equations to $\mathcal{E}$. \Cref{example:overall} provides a more
detailed explanation of \cref{alg:compilewithbasecases}.

\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Compile. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we obtain the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, resulting in $g(l, 0) = 1$.
\end{example}

It is worth noting that these base cases overlap when $l = m = 0$ but remain
consistent since $0^{0} = 1$. Generally, let $\phi$ be a formula with two
domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the WFOMC
of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the
WFOMC of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

Finally, we highlight the crucial role of the \Simplify procedure in simplifying
the algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here:
\begin{enumerate*}[label=(\roman*)]
  \item $n$ is a variable,
  \item $a, b \in \mathbb{N}_{0}$ are constants,
  \item $f$ is an expression that may depend on $m$, and
  \item $[a \le m \le b] =
  \begin{cases}
    1 & \text{if $a \le m \le b$} \\
    0 & \text{otherwise}
  \end{cases}$ represents the Iverson bracket.
\end{enumerate*}
\Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overall}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into the
simplified form mentioned above.

@@

\Cref{alg:findbasecases} summarises the implementation of the \FindBaseCases
function. When a function $f$ calls itself recursively, \FindBaseCases considers
two types of arguments:
\begin{enumerate*}[label=(\roman*)]
  \item constants and
  \item arguments of the form $x_{i} - c_{i}$, where $c_{i}$ is a constant and
  $x_{i}$ is the $i$-th argument of the signature of $f$.
\end{enumerate*}
If the argument is a constant $c_{i}$, a base case with $c_{i}$ is added. In the
second case, a base case is added for each constant from zero up to (but not
including) $c_{i}$. The following discussion explains the reasoning behind this
approach.

\begin{example}
  Let us consider the recursive function $g$ from \cref{example:overall}. In
  this case, \FindBaseCases iterates over two function calls: $g(l-1, m)$ and
  $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the latter
  produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

For the rest of this section, let $\mathcal{E}$ represent the equations returned
by \CompileWithBaseCases. To demonstrate that the base cases identified by
\FindBaseCases are sufficient, we begin with a few observations that stem from
the details of previous
work~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:conf/kr/DilkasB23} and this work.

\begin{fact}\label{assumption1}
  For each function $f$, there is precisely one equation $e \in \mathcal{E}$
  with $f(\mathbf{x})$ on the left-hand side where all $x_{i}$'s are variables
  (i.e., $e$ is not a base case). We refer to $e$ as the \emph{definition} of
  $f$.
\end{fact}

\begin{fact}\label{assumption2}
  There is a \emph{topological ordering} of all functions ${(f_{i})}_{i}$ in
  $\mathcal{E}$ such that equations in $\mathcal{E}$ with $f_{i}$ on the
  left-hand side do not contain function calls to $f_{j}$ with $j > i$. This
  condition prevents mutual recursion and other cyclic scenarios.
\end{fact}

\begin{fact}\label{assumption3}
  For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, the evaluation
  of $\expr$ terminates when provided with the values of all relevant function
  calls.
\end{fact}

\begin{corollary}\label{fact}
  If $f$ is a non-recursive function with no function calls on the right-hand
  side of its definition, then the evaluation of any function call
  $f(\mathbf{x})$ terminates.
\end{corollary}

\begin{fact}\label{fact2}
  For any equation $f(\mathbf{x}) = \expr{}$, if $\mathbf{x}$ contains only
  constants, then $\expr{}$ cannot include any function calls to $f$.
\end{fact}

Additionally, we introduce an assumption about the structure of recursion.

\begin{assumption}\label{assumption4}
  For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every recursive
  function call $f(\mathbf{y}) \in \expr$ satisfies the
  following:
  \begin{itemize}
    \item Each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some constant
          $c_{i}$.
    \item There exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
          $c_{i} > 0$.
  \end{itemize}
\end{assumption}

Finally, we assume a particular order of evaluation for function calls using the
equations in $\mathcal{E}$. Specifically, we assume that base cases are
considered before the recursive definition. The exact order in which base cases
are considered is not relevant.

\begin{assumption}
  When multiple equations in $\mathcal{E}$ match a function call
  $f(\mathbf{x})$, preference is given to an equation with the most constants on
  its left-hand side.
\end{assumption}

With the facts and assumptions mentioned above, we prove the following theorem.

\begin{theorem}[Termination]\label{thm:halting}
  Let $f$ be an $n$-ary function in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbf{N}_{0}^{n}$. Then the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}
